---
title: 'Geog6300: Lab 6'
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Regression

```{r setup, message=FALSE, warning=FALSE}
library(sf)
library(tidyverse)
library(tmap)
library(ggplot2)
library(broom)
```

**Overview:**
This lab focuses on regression techniques. You'll be analyzing the association of various physical and climatological characteristics in Australia with observations of several animals recorded on the citizen science app iNaturalist.

###Data and research questions###

Let's import the dataset.

```{r}
lab6_data<-st_read("data/aus_climate_inat.gpkg")

tm_shape(lab6_data)+tm_polygons("ndvi_20_med")
```

The dataset for this lab is a 1 decimal degree hexagon grid that has aggregate statistics for a number of variables:

* ndvi: NDVI/vegetation index values from Landsat data (via Google Earth Engine). These values range from -1 to 1, with higher values indicating more vegetation.
* maxtemp_00/20_med: Median maximum temperature (C) in 2000 or 2020 (data from SILO/Queensland government)
* mintemp_00/20_med: Median minimum temperature (C) in 2020 or 2020 (data from SILO/Queensland government)
* rain_00/20_sum: Total rainfall (mm) in 2000 or 2020 (data from SILO/Queensland government)
* pop_00/20: Total population in 2000 or 2020 (data from NASA's Gridded Population of the World)
* water_00/20_pct: Percentage of land covered by water at some point during the year in 2000 or 2020
* elev_med: Median elevation (meters) (data from the Shuttle Radar Topography Mission/NASA)

There are also observation counts from iNaturalist for several distinctively Australian animal species: the central bearded dragon, the common emu, the red kangaroo, the agile wallaby, the laughing kookaburra, the wombat, the koala, and the platypus.

Our primary research question is how the climatological/physical variables in our dataset are predictive of the NDVI value. We will build models for 2020 as well as the change from 2000 to 2020. The second is referred to as a "first difference" model and can sometimes be more useful for identifying causal mechanisms.

###Part 1: Analysis of 2020 data###

We will start by looking at data for 2020. 

**Question 1** _Create histograms for NDVI, max temp., min temp., rain, and population, and water in 2020 as well as elevation. Based on these graphs, assess the normality of these variables._

```{r}

variables <-c("ndvi_20_med", "maxtemp_20_med", "mintemp_20_med", "rain_20_sum", "pop_20", "water_20_pct", "elev_med")

for (var in variables) {
  hist(lab6_data[[var]], main = paste("Histogram of", var), xlab = var, col = "grey", breaks = 30)}


```

{Visually, all of these graphs appear to be skewed and not normal. Elevation is probably the closest to normal distribution. }

**Question 2** _Use tmap to map these same variables using Jenks natural breaks as the classification method. For an extra challenge, use `tmap_arrange` to plot all maps in a single figure._

```{r}
tmap_mode("plot")

tm_list <- lapply(variables, function(var) {
  tm_shape(lab6_data)+
    tm_polygons(var,style = "jenks", palette = "viridis", title = var)+
    tm_layout(legend.outside = TRUE)})

tmap_arrange(plotlist = tm_list)

```


**Question 3** _Based on the maps from question 3, summarise major patterns you see in the spatial distribution of these data from any of your variables of interest. How do they appear to be associated with the NDVI variable?_

{The ndvi(vegetation) is highest around the edges of australia, primarily on the southeast coast. This appears to have a similar pattern as the rain map and population. The temperature has a distinct pattern of warmer to colder from the north to south, respectively.}

**Question 4** _Create univariate models for each of the variables listed in question 1, with NDVI in 2020 as the dependent variable. Print a summary of each model. Write a summary of those results that indicates the direction, magnitude, and significance for each model coefficient._

```{r}
result<-lapply(variables, function(var){
  model<-lm(ndvi_20_med ~ lab6_data [[var]], data = lab6_data)
  summary(model)})

print(result)
```

{*REVISED:  
2)Max temp:For one degree max temp increase, there is a slight negative effect (-0.013) on the vegetation index. This was a signifcant value (p<0.0001) and the r-squared was 0.21. 
3)Min temp:For one degree min temp increase, there is also a similar slight negative effect (-0.012) on the vegetation index. This was a significant value (p<0.0001) and the r-squared was 0.14.
4)Rainfall: As rain fall went up by 1mm, there was a positive effect on vegetation index (9.12e-7). This was a signifcant value (p<0.0001) and the r-squared was 0.43 indicating a relatively strong relationship. 
5)Population:As population went up by one, the vegetation index would increase by 1.5e-6. This was a signifcant value (p<0.0001) and the r-squared was 0.13.
6)% of land covered by water: As percent of land covered by water went up by 1, the vegetation index went down 0.18. This was not a signifcant result (p>0.05) adn the r-squared was minimal at 0.002.
7)Median elevation: As median elevation went up by 1 meter, the vegation index increased by 1.787e-4. This was a signifcant value (p<0.0001) and the r-squared was small at 0.05. 


*original: All had significant findings except for water_20_pct (which also had the lowest r squared value). rain_20_sum had the highest r-squared value at 0.43. Max temp, mintemp, and water_pct have a negative effect on vegetation. Rain, pop, and elevation had positive effect on vegetation. }

**Question 5** _Create a multivariate regression model with the variables of interest, choosing EITHER max or min temperature (but not both) You may also choose to leave out any variables that were insignificant in Q4. Use the univariate models as your guide. Call the results._

```{r}
multi_reg_model<-lm(ndvi_20_med ~ maxtemp_20_med + rain_20_sum + pop_20 + elev_med,data = lab6_data)

summary(multi_reg_model)
```

**Question 6** _Summarize the results of the multivariate model. What are the direction, magnitude, and significance of each coefficient? How did it change from the univariate models you created in Q4 (if at all)? What do the R2 and F-statistic values tell you about overall model fit?_

{*REVISED: 
1) Max temp: Signifcant (p<0.05); As max temp increases 1, vegetation index decreases by 1.1e-2. This is a smaller value than in Q4. 
2)Rainfall: Signifcant (p<0.05); As rainfall increased by 1mm, vegetation index increased by 8.46e-7. This is slightly smaller but almost the exact same as Q4.
3)Population:Signifcant (p<0.05); As population increased by 1, vegetation index increased by 2.86e-7. This decreased from Q4.
4)Median Elevation:Signifcant (p<0.05); As median elevation increased by 1 meter, vegetation index increased by 1.23e-4. This is slightly smaller than Q4. 
The r-squared value of 0.64 shows that this is a relatively strong relationship, compared to the individual coefficients in Q4. The high f-statistic of 315.5 and p-value < 2.2e-16 tells us this model is a good predictor of vegetation index. 


*original: All variables are statistically significant. maxtemp is the only negative relationship. There is a relatively high r-squared value of 0.64 which is much higher than any of the individual variables in q4. The high r-squared and f-statistic values tell us that this is a good overall model fit. }

**Question 7** _Use a histogram and a map to assess the normality of residuals and any spatial autocorrelation. Summarise any notable patterns that you see._

```{r}
lab6_data$residuals <-residuals(multi_reg_model)

residuals <- as.numeric(residuals(multi_reg_model))

hist(residuals, main = "Histogram of Residuals", xlab = "Residuals", col = "lightgrey",breaks = 30)

tm_shape(lab6_data)+
  tm_polygons("residuals", style = "jenks", palette = "RdBu",  midpoint = 0, title = "Residuals") +
  tm_layout(legend.outside = TRUE)

tmap_mode("view")
```

{The residuals are relatively evenly spreadout throughout. The red (lowest) residuals are found along the north and south coast, with few on the east and west coast. The orange residuals seem to be clustered around the middle region.}

**Question 8** _Assess any issues with multicollinearity or heteroskedastity in this model using the techniques shown in class. Run the appropriate tests and explain what their results show you._

```{r}
library(car)

vif(multi_reg_model)
    
ncvTest(multi_reg_model)
```

{The vif values are from 1.009-1.17 indicating no multicollinearity. The statisically significant pvalue indicates evidence of heterskedasity. }

**Question 9** _How would you summarise the results of this model in a sentence or two? In addition, looking at the full model and your diagnostics, do you feel this is a model that provides meaningful results? Explain your answer._

```
{Maxtemp, rain, pop, and elevation have an effect on ndvi. I feel this model provides meaningful results due to the statistical significance and the relatively high r-squared value. However, the heteroskedacity means this model may need revisiting. }
```


**Disclosure of assistance:** _Besides class materials, what other sources of assistance did you use while completing this lab? These can include input from classmates, relevant material identified through web searches (e.g., Stack Overflow), or assistance from ChatGPT or other AI tools. How did these sources support your own learning in completing this lab?_

{chatgpt helps me diagnose when an error occurs such as not having a certain package running (library()). }

**Lab reflection:** _How do you feel about the work you did on this lab? Was it easy, moderate, or hard? What were the biggest things you learned by completing it?_

{I felt moderate about this lab. The class scripts and chatgpt helped with completing code. The in class slides helped with interpreting results. }


**Challenge question**

#Option 1
Create a first difference model. To do that, subtract the values in 2000 from the values in 2020 for each variable for which that is appropriate. Then create a new model similar to the one you created in question 5, but using these new variables showing the *change in values* over time. Call the results of the model, and interpret the results in the same ways you did above. Also chart and map the residuals to assess model error. Finally, write a short section that summarises what, if anything, this model tells you. 

```{r}
#chose option 2
```




#Option 2
The animal data included in this dataset is an example of count data, and usually we would use a Poisson or similar model for that purpose. Let's try it with regular OLS regression though. Create two regression models to assess how the counts of two different animals (say, koalas and emus) are associated with at least three of the environmental/climatological variables given above. Be sure to use the same independent variables in each model. Interpret the results of each model and then explain the importance of any differences in the model coefficients between them, focusing on direction, magnitude, and significance.

```{r}
koala_model<-lm(Koala ~ maxtemp_20_med + rain_20_sum +pop_20,data = lab6_data)

emu<-lm(Common.emu ~ maxtemp_20_med + rain_20_sum +pop_20,data = lab6_data)

summary(koala_model)
summary(emu)
```
{For koala model, the r-squared is 0.11, with an f-statistic of 29.28, and p-value<0.0001 indicating a good overall fit. There is a positive correlation between the three variables (max temp = 7.5e-1, rainfall = 9.6e-5, population = 2.6e-3). Only population has a statiscally significant p-value (p<0.0001). 

For emu model, the f-statistic 27.78 and p-value <0.0001 indicate a good overall fit. Similar to koalas, the r-squared value was 0.10. Max temp and population had signifcant p-values (p<0.001). Max temp had a negative correlation (-2.1) while the other two were positive (rainfall = 1.6e-6, population = 1.27e-4). 

Both animal models had similarly low R-squared values. Both animals had a dependence on population (higher effect on koalas) but only emus had a statisically signifcant relationship with max temp. Rainfall was not statistically significant for either animal. }

library(knitr)

